Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2018-02-24 paper.tex
new images
images of 4l2l solution + new  table

The reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called a tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with the amino acids surrounding the protein active site~\cite{gora2013gates} (see Fig.~\ref{fig::motiv}).
The location and size of a tunnel determine the size and shape of a ligand passing through this tunnel to the active site.
Currently, the biochemists can use various computational methods for tunnel detection in order to estimate the possibility of interactions between the protein and ligand.
This can help them to reduce the number of necessary in-vitro experiments.
The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), 
i.e., they approximate the ligand by a bounding sphere.
However, ligands are typically of a non-spherical shape, which is completely ignored by these methods.

@@

In this paper, we propose a modification of the RRT planner to find trajectories for non-spherical flexible ligands inside protein tunnels.
Random samples are generated only around a given tunnel instead of the whole configuration space.
%We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
The potential tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
To enable the movement of the ligand in the narrow parts of the tunnel, the free-space is dilated by shrinking the atom radii, similarly, e.g., to~\cite{hsu06multilevel}.  %cortes2005path removed as it is not clear it they really rescale
The ligand flexibility is modeled using a fixed set of predefined conformations which can be taken by the ligand.

@@

\section{Related Work}
Transportation of ligands into the protein active site can be directly evaluated using molecular dynamics (MD) simulations, where both the protein and ligand are flexible.
However, MD simulations are considerably computational demanding~\cite{kingsley2014including} and they have to be prepared
for a particular ligand/protein pair, which is not practical for testing a large number of different ligands interacting with the protein.
Therefore, computational approaches that consider only the static geometry of the ligand are still preferable due to their high speed.

@@

The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using the sampling-based motion planners.
These planners can cope with objects (robots) with many Degrees of Freedom (DOF) and they can also consider objects of an arbitrary shape.
Two widely used planners are the Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and the Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
%Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
%The computation of the ligand trajectories can be solved also using motion 
%In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
%In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
%A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
%If $\qnew$ is collision-free, it is added to the tree.
%The algorithm terminates if the tree approaches the goal configuration close enough.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
The well known issue of the sampling-based planners is the narrow passage problem.
The narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
Due to its low volume, the probability of sampling the narrow passage is low.
Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
The PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g., by sampling along the medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
In~\cite{bergWIG}, the workspace is represented by a grid in which the difficult regions are identified using watershed labeling.
The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge} or even identified by a human operator~\cite{denny2018general}. 
%The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
%based only on the features of the workspace~\cite{hannaWIS}.
%.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
%~\cite{hsuBridge,overmarsGauss}.

@@

The RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
To prevent this blocking, the DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
An automatic adaptation of this radius was proposed in the ADD-RRT~\cite{jailletADRRT}.
%The disadvantage of DD-RRT is the sensitivity to the predefined radius.
%Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
The guiding path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
It can be expected that the tunnels in MD simulations without ligands are rather narrow.
However, the tunnels can adapt their shape to the passing ligands and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can potentially serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
In our solution, we simulate this by shrinking the atom radii of the ligand, which is used also by~\cite{cortes2010simulating,guieysse2008structure}.

Molecular dynamics simulations of ligands passing through proteins show that the ligands can be highly flexible.
However, ligands cannot form arbitrary conformations. 
The ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
the dimension of the configuration space is not increased as in~\cite{cortes2010simulating}.
%So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
%The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.

@@

The second modification is the expansion procedure (Alg.~\ref{alg::expand}). % which expands the tree from $\qnear$ towards $\qrand$.
%The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
%The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
%$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in the narrow void space.
We expand the tree considering all $\L$ conformations.
%The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with the highest scale.
First, the maximal scale $\smax \in \S$ is considered and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales, and the
smaller scales are used only in difficult (narrow) regions.

@@

In the case of large sets $\L$, only a subset of the available ligand conformations can be used in the planning.
To select the subset, the conformations are ranked using the spread of the atoms around the geometric center of the ligand.
The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
distances then characterizes the spread.
Low deviation is characteristic for `packed' conformations, while larger deviations correspond to the elongated conformations (Tab.~\ref{fig::m040c}).
To select the subset of $n$ representative conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.
The selected conformations are then used in the planning process.

@@

The proposed algorithm was verified in two scenarios: on a sequence frames of MD simulation and on a set of protein conformations provided by the tCONCOORD tool~\cite{seeliger2007geometry}.

@@

Models  of the ligands in Mol2 formats were converted to the param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.
For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
For each initial configuration, 10 trajectories were computed, which resulted
in $100 \times 4 \times 20 \times 10 = 80,000$ trajectories per ligand.

@@

The methods \RA\ and \RB\ differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
The retraction step is repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per ligand conformation, while \RA\ extends the tree only by one new node.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
For each conformation and tested scale, the \RA\ method attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
Other parameters of the \RB\ planner were the same as for \RA.
The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 2~\AA\ (i.e., the 
geometric center of the ligand was closer than 2~\AA\ from the active site).

@@

The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand's 
name shows the number of the selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denotes the number of atoms of the ligand.
In case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
For the DCP ligand, the higher traversability rate was achieved when 50 conformations were selected.
This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using the smaller number of conformations or when preferring packed conformations.

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
    The tested protein consists of 4650 atoms. 
    The rate is shown in [\%].
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
%\renewcommand{\arraystretch}{0.7}
\input graphSelNew.tex
}
\end{table}

@@

Based on the results of the previous experiment, 50 conformations using the `S' strategy were selected for ligands 
m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane).

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
    The tested protein consists of 4650 atoms.
    The rate is shown in [\%].
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
%\renewcommand{\arraystretch}{0.7}
%\input graphNew.tex %with \rc (a1*) column
\input graphNewNoA3.tex %without \rc (a1*) column
}
\end{table}

The traversability rate is shown in Tab.~\ref{tab::main}. 
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (consisting of 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the DCP ligand, the trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
However, their binding was not yet approved by dedicated MD simulations.

Besides scaling-down the ligand, also the atoms of the protein can be scaled, which is used, e.g., in~\cite{cortes2005path}.
The tool~\cite{cortes2005path} uses the scale 0.8 for both the ligand and protein.
We tested also this option using \RA\ (the column \RD\ in Tab.~\ref{tab::main}).
The results are similar to the results of \RA\ and $\smin=0.6$ (except for DCP and m040, which can be a result of randomization).

@@

\begin{table}[t]
\caption{\label{fig::m040c}
\small
Examples of elongated/packed conformation for m040 (1,5-Dichloropentane).
The table shows the average number of conformations (out of 50) used in trajectories reaching the active site.
}
\centering
{\footnotesize
%\def\arraystretch{0.9}
\begin{tabular}{cc}
\rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.1\textwidth]{fig/m040-conf1} \\
\rotatebox{0}{Packed} \\ \includegraphics[width=0.078\textwidth]{fig/m040-conf2}
\end{tabular}
}
{\scriptsize
\def\arraystretch{0.9}
\input usedc.tex
}
\end{table}

@@

%The case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
The task of the second analysis was to find the trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
site of A4 hydrolase (PDB ID 4L2L, 9666 atoms).
%This ligand is a known inhibitor of the protein 4L2L.
%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L (without the inhibitor).
The initial PDB structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
The protein conformations were generated using the default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
The tunnels leading to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0, using the probe 0.9~\AA.
All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
The ordering of the detected tunnels slightly differ in different conformations.
Therefore, the ligand traversability towards the active site was computed for the first three widest tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
The conformations of the ligand were prepared in the same way as in the previous section.
From the available $|\L|=736$ conformations, 100 were selected using the `S' strategy.

As the tested ligand is a known inhibitor of the protein 4L2L, we assume that the ligand can reach the active site.
We tested several scaling-down factors from $\smin=0.4$ to $\smin=0.6$.

The example of the found trajectory showing the final position of the ligand is depicted in Fig.~\ref{fig::inhibitor}.
TODO: report num of traversable frames

>>>>

The reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called a tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with the amino acids surrounding the protein active site~\cite{gora2013gates,marques2017enzyme} (see Fig.~\ref{fig::motiv}).
The location and size of a tunnel determine the size and shape of a ligand passing through this tunnel to the active site.
Currently, the biochemists can use various computational methods for tunnel detection in order to estimate the possibility of interactions between the protein and ligand.
This can help them to reduce the number of necessary in-vitro experiments.
The most common tools detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), 
i.e., they approximate the ligand by a bounding sphere.
However, ligands are typically of a non-spherical shape, which is completely ignored by these methods.

@@

In this paper, we propose a modification of the RRT planner to find trajectories for non-spherical flexible ligands inside protein tunnels.
A tunnel is represented by a set of consecutive spheres surrounding the tunnel centerline.
Random samples are generated only around a given tunnel instead of the whole configuration space.
%We localize the generation of the random samples, i.e., instead of sampling the whole configuration space we focus only on the space surrounding a potential tunnel detected by the Voronoi-based method (using a smaller ligand probe in order to avoid omitting possibly feasible tunnel candidates).
To enable the movement of the ligand in the narrow parts of the tunnel, the free-space is dilated by shrinking the atom radii, similarly, e.g., to~\cite{hsu06multilevel}.  %cortes2005path removed as it is not clear it they really rescale
The ligand flexibility is modeled using a fixed set of predefined conformations which can be taken by the ligand.

@@

\section{Related Work}
Transportation of ligands into the protein active site can be directly evaluated using molecular dynamics (MD) simulations, where both the protein and ligand are flexible.
However, MD simulations are considerably computational demanding~\cite{kingsley2014including} and they have to be prepared
for a particular ligand/protein pair, which is not practical for testing a large number of different ligands interacting with the protein.
Therefore, computational approaches considering only geometry of ligand and proteins are still preferable due to their high speed.

@@

The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using the sampling-based motion planners.
These planners can cope with objects (robots) with many Degrees of Freedom (DOF) and they can also consider objects of an arbitrary shape.
Two widely used planners are the Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and the Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
%Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
%The computation of the ligand trajectories can be solved also using motion 
%In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
%In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
%A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
%If $\qnew$ is collision-free, it is added to the tree.
%The algorithm terminates if the tree approaches the goal configuration close enough.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
The well known issue of the sampling-based planners is the narrow passage problem.
The narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
Due to its low volume, the probability of sampling the narrow passage is low.
Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
The PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g., by sampling along the medial axis~\cite{wilmarthMAPRM}.
In~\cite{bergWIG}, the workspace is represented by a grid in which the difficult regions are identified using watershed labeling.
The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge} or even identified by a human operator~\cite{denny2018general}. 
%The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
%based only on the features of the workspace~\cite{hannaWIS}.
%.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
%~\cite{hsuBridge,overmarsGauss}.

@@

The RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in the difficult regions, as the growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for the expansion to a small ball. 
The radius of this ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
An automatic adaptation of this radius was proposed in the ADD-RRT~\cite{jailletADRRT}.
%The disadvantage of DD-RRT is the sensitivity to the predefined radius.
%Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying how to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where the whole path is used to generate samples in the configuration space.
The guiding path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

It has been observed, that the tunnels detected in protein conformations without ligands (provided by MD simulations without ligands or by
protein conformation sampling) are rather narrow. 
However, in protein dynamics with ligand, the tunnels can move, merge and adapt to the shape of the passing ligand and vice versa.
%Therefore, it can be expected that the detected tunnels are rather narrow.
%Due to the protein dynamics, the tunnels move and merge, and their bottleneck can change in each time step.
%The tunnels can adapt their shape to the passing ligands and vice versa.
In consequence, even very narrow tunnels, being too small for the passage of a given ligand-approximating sphere, can potentially serve as the transportation path for the ligand.
Therefore, also the existing Voronoi-based methods enable the user to decrease the size of the approximating sphere which simulates the shrinking of the ligand size.
In our solution, we simulate this by shrinking the atom radii of the ligand, which is used also by~\cite{cortes2010simulating,guieysse2008structure}.

MD simulations of ligands passing through proteins further show that the ligands can be highly flexible.
The possible conformations of a ligand are determined by the torsion and dihedral angles between its atoms. 
The low-energy conformations are preferred.
%the ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
Therefore, the ligand flexibility can be modeled using only a predefined set of low-energy conformations.
By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
the dimension of the configuration space is not increased as in~\cite{cortes2010simulating}.
%So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
%The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.

@@

The second modification is the expansion procedure (Alg.~\ref{alg::expand}). % which expands the tree from $\qnear$ towards $\qrand$.
%The expansion towards $\qrand$ is crucial for the growth of the tree towards the unexplored areas of the configuration space.
%The traditional straight-line expansion (i.e., new configuration is constructed on the line from $\qnear$ to $\qrand$ in the distance
%$\varepsilon$ from $\qnear$) more likely leads to collisions, especially in the narrow void space.
We expand the tree considering all $\L$ conformations.
%The proposed expansion attempts to expand the tree using all the conformations $\L$ and by preferring larger scales than the smaller ones.
For each $l \in \L$, the expansion procedure attempts to find a new collision-free configuration around $\qnear$ with the highest scale.
First, the maximal scale $\smax \in \S$ is considered and $m$ random samples are generated around $\qnear$ (in the distance $\varepsilon$) 
and tested for collisions.
The random samples are generated similarly as in the case of $\qrand$ samples, the translation part $(x,y,z)$ is generated around $\qnear$.
The nearest collision-free sample towards $\qrand$ is selected and added to the tree.
If none of the tested samples is collision-free, the scale is reduced to $\smax-\sdelta$ and the search continues
until a collision-free sample is found or until the minimal reduced-scale $\smin$ is reached.
By testing first the largest scale ($\smax$), the expansion procedure prefers expanding the tree by larger scales, and the
smaller scales are used only in difficult (narrow) regions.

@@

In the case of large sets $\L$, only a subset of the available ligand conformations can be used in the planning.
To select the subset, the conformations are ranked using the spread of the atoms around the geometric center of the ligand.
The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
distances then characterizes the spread.
Low deviation is characteristic for `packed' conformations, while larger deviations correspond to the elongated conformations (Tab.~\ref{tab::m040c}).
To select the subset of $n$ representative conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.
The selected conformations are then used in the planning process.

@@

The proposed algorithm was verified in two scenarios: on a sequence of frames of MD simulation and on a set of protein conformations provided 
the conformation sampling approach.
%by the tCONCOORD tool~\cite{seeliger2007geometry}.

@@

Models  of the ligands in Mol2 formats were converted to the param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
Four different scaling-down factors were used: $\S=\{0.5,0.6,0.7,0.8\}$.

For each scale, $|\QI|=20$ collision-free initial configurations were found in the $\RI=2$~\AA\ radius around the first sphere of the tunnel (outside the protein).
For each initial configuration, 10 trajectories were computed, which resulted
in $100 \times 4 \times 20 \times 10 = 80,000$ trajectories per ligand.

@@

The methods \RA\ and \RB\ differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
The retraction step is repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per ligand conformation, while \RA\ extends the tree only by one new node.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
For each conformation and tested scale, the \RA\ method checks $m=100$ candidate configurations for collision, and \RB\ checks $5 \times 20$ configurations for collision.
%collision-detection queries.
Other parameters of the \RB\ planner were the same as for \RA.
The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 2~\AA\ (i.e., the 
geometric center of the ligand was closer than 2~\AA\ from the active site).

@@

The traversability rates are shown in Tab.~\ref{tab::selection}, where the subscript of each ligand's 
name shows the number of the selected conformations (5 or 50) and the used method (S or P).
The number after the slash in the ligand's name denotes the number of atoms of the ligand.
In case of both tested ligands and 50 selected conformations, the `S' strategy leads to a higher traversability rate than the `P' strategy.
For the DCP ligand, the higher traversability rate was achieved when 50 conformations were selected.
This experiment shows that selecting 50 conformations using the `S' strategy (Sec.~\ref{sec::strat}) leads to better results
than using the smaller number of conformations or when preferring the packed conformations.

\begin{table}[bt]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
    The rate is shown in [\%].
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
%\renewcommand{\arraystretch}{0.7}
\input graphSelNew.tex
}
\end{table}

@@

Based on the results of the previous experiment, 50 conformations using the `S' strategy (Sec.~\ref{sec::strat})  were selected for ligands 
m037t (1,2-Dichloroethane), m038t (1,3-Dichloropropane), m056 (2-Chlorobutane), and m080 (Trichloropropane).

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
%\renewcommand{\arraystretch}{0.7}
%\input graphNew.tex %with \rc (a1*) column
\input graphNewNoA3.tex %without \rc (a1*) column
%\input traversability.tex
}
\end{table}

The traversability rate is shown in Tab.~\ref{tab::main}. 
Only the traversability of the main tunnel is shown, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (consisting of 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the DCP ligand, the trajectories were detected in 18~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
However, their binding was not yet approved by dedicated MD simulations.
Most of the used 50 ligand conformations where used in the planning, which is shown in Tab.~\ref{tab::m040c}.

Besides scaling-down the ligand, also the atoms of the protein can be scaled, which is used, e.g., in~\cite{cortes2005path}, where
%The tool~\cite{cortes2005path} uses 
the scale 0.8 is used for both the ligand and protein.
We tested also this option using \RA\ (the column \RD\ in Tab.~\ref{tab::main}).
The results are similar to the results of \RA\ and $\smin=0.6$ (except for DCP and m040, which can be a result of randomization).

@@

\begin{table}[t]
\caption{\label{tab::m040c}
\small
Examples of elongated/packed conformation for m040 (1,5-Dichloropentane).
The table shows the average number of conformations (out of 50) used in trajectories reaching the active site.
}
\centering
{\footnotesize
%\def\arraystretch{0.9}
\begin{tabular}{cc}
\rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.07\textwidth]{fig/m040-conf1} \\
\rotatebox{0}{Packed} \\ \includegraphics[width=0.0375\textwidth]{fig/m040-conf2}
\end{tabular}
}
{\scriptsize
\def\arraystretch{0.9}
\input usedc.tex
}
\end{table}

@@

%The case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
The task of the second analysis was to find the trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine into the active
site of A4 hydrolase (PDB ID 4L2L, 9666 atoms).
%This ligand is a known inhibitor of the protein 4L2L.
%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L.
The initial PDB structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
The protein conformations were generated using the default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
The tunnels leading to the active site defined by residues 134 (Alanine) and 311 (Phenylalanine) were computed by CAVER 3.0, using the probe 0.9~\AA.
All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
The ordering of the detected tunnels slightly differ in different conformations.
Therefore, the ligand traversability towards the active site was computed for the first three widest tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA\, and 1.19~\AA\ with the average lengths 22.3~\AA, 25.3~\AA\, and 28.4~\AA, respectively).
The conformations of the ligand were prepared in the same way as in the previous section.
From the available $|\L|=736$ low-energy conformations, 100 were selected using the `S' strategy.

\begin{table}[bt]
\centering
\caption{\label{tab::4l2l}
    \small
    Traversability rate [\%] in 1000 conformations of the protein 4L2L in the first three tunnels.
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
\input traversability.tex
}
\end{table}
